Blocks of extended ancestry around selected loci in hybrid zones

### 2016-02-28 drafts/hybrid_zone.tex
edits

\alisa{A conceptual cartoon of this process is shown in Figure~\ref{Fig:schematic}.}
\plr{Connect these up more.}

@@

The simulations were executed in R, with scripts available on
\url{https://github.com/asedghifar/hybridzones}.
\plr{if I'm to be the corresp author, put this to \url{https://github.com/petrelharp/clinal-lineages}?}

@@

With the above theory and simulations, we are able to make predictions about patterns surrounding selected loci. In reality, however, such loci are not known, so it is useful to have per-site statistics that may allow for detection of candidate targets of selection. The most straightforward measure is $l_I(m)$ the length of haplotype chunk that a sampled chunk containing genomic position $b$, conditioned on ancestry $I$ at the site. Based on our expectation of the distribution of block lengths, we compute $l_B(m)$ for $m$ across a chromosome.

%Another statistic is the cumulative probability $C_I(m)$ of a sampled chunk encompassing position $m$ being longer than a chunk sampled from the distribution of all chunk lengths across the genome in the population.

Additionally, we look at the mean value of these statistics in the two chunks, $m-$ and $m+$, that flank the block of unbroken ancestry containing $m$. Since there appears to be a signal at these regions also, we further define a statistic

$$C =  \frac{2\sum{l_B(k_i)}}{\sum{l_A(k_i+)+l_A(k_i-)}}$$

which is the ratio of mean block length around $k_i$ and mean length of blocks directly flanking the block surrounding $k_i$.

@@

As expected, regions surrounding a locus under selection are more resistant to introgression,
as seen in figures~\ref{Fig:resistanceToIntrogression1000g} and \ref{Fig:resistanceToIntrogression100g}).
When present, however, we expect haplotypes including the locally less common allele
to be longer than the genome-wide average.
Indeed,
as shown in Figures \ref{Fig:blockLengths} and \ref{Supp:blockLengthHeatmap},
the mean length of such haplotype blocks are up to three times longer
than the average for that geographic location.
\plr{Write down in the section on statistics what exactly you normalize by:
is it the mean for that geographic location?  Or, what?}
% This pattern provides a potential avenue for identifying targets of selection by looking for a signal of increased $l_B$ or $l_A$ for positions surrounding a selected site, compared to the genome-wide background  at geographic sites far from the center of the zone (Figure~\ref{Fig:blockLengths},~\ref{Supp:blockLengthHeatmap}). 
The mean haplotype length found without conditioning on ancestry ($l_\cdot$)
also shows an small increase near the selected locus,
but it is much smaller (see Figures \ref{Supp:blockLengthHeatmapNoAnc} and \ref{Supp:blockLengthNoAnc}).

@@

\subsection*{Lengths of adjacent blocks}

\plr{Define this statistic and put in the Summary Stats section. Also, move this up with the haplotype length stuff?}
We find that $l_A(k\pm)$, the length of blocks adjacent to blocks containing $k$, are shorter than average when $k$ is in a block of unbroken ancestry surrounding a selected site on the ``wrong side'' of the side of the zone.  (Figure~\ref{Supp:adjacentBlocks}). This pattern provides an avenue for further enhancement of the block-length signal around a selected locus, by taking the ratio $\frac{2\sum{l_B(k_i)}}{\sum{l_A(k_i+)+l_A(k_i-)}}$ of mean block lengths surrounding a site with the mean length of adjacent blocks (Figures~\ref{Supp:ratioBlockAdjacent},~\ref{Supp:ratioBlockAdjacentHeatmap}).

@@

\begin{figure}
\includegraphics{figs/blocksAlongChromAncBConditioning}
\caption{Mean block length, $l_B(k_i)$, along a simulated chromosome of length $1M$ with selected locus at position $0.5M$. Here each line represents  $l_B(\mathbf{k})$, the mean block lengths in a given deme, and is normalized by mean block length $l_B(k_i)$ along the chromosome in the deme.
    \plr{what is $T$ and $s$? also, add version with different $T$ and/or $s$ to the supplement.}
}\label{Fig:blockLengths}
\end{figure}

@@

\plr{This needs a bit more precise discussion.}

Although our analytical model does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $T$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family. We find that, as distance from the zone center increases, the average family size of an ancestor from time $T$ decreases (Figure~\ref{Fig:family_size}). One interpretation of this observation is that migrants tend to leave few offspring before being removed from the population by selection. This can potentially provide additional signal for characterizing lineages strongly influenced by selection.

@@

Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models (e.g., ecological selection \citep{Barton1989,Barton1993}). 
We expect, therefore, our findings here to be somewhat generalizable.

@@

@@

\plr{This next stuff is said later, under "Assumptions"; maybe should be moved there.}
\yb{While our theory focusses on a spatially arranged hybrid zone, it is likely that our results extend to alternative models of hybridization including models of assortative mating and hybrid swarms. This is because regardless of the specific demographic model, selection against hybrids will severely limit opportunities for recombination between ancestries around incompatibilities.}

@@

\plr{(1D)}
For ease of calculations, our analytic model assumes a continuous and smooth one-dimensional spatial structure of the geographic landscape. This allows us to derive analytic expressions that track lineages, and is an example of the more general idea of duality for particle systems \alisa{I think this may be beyond the average ME reader?}. 
\plr{Tracking lineages is duality, has nothing to do with 1D, though.}
However, this assumes a large population density limit that excludes \plr{coalescence} and makes linages Markov \alisa{not sure about what point this is making? Is there a better way to state this for our intended audience?}.  
\plr{It means we're ignoring drift/coalescence, an important source of stochasticity; but that's the next paragraph too}
In contrast, our simulations model regularly spaced demes. In reality, populations may be patchily connected, especially at the edges of species ranges where hybrid zones may occur. While we still believe that the genomic signature of long ancestry chunks around selected sites will still hold in these situations, the strength of this expectation is unclear.

Lastly, the analytic model presented here lacks \yb{both coalescence and a true pedigree structure}. \alisa{This can result in an underestimate of covariance in ancestry, as we have ignored sharing through genealogy \cite{Liang2014}.} 
The broad agreement between our analytic predictions and simulations that include an explicit pedigree structure suggests that this assumption does not have a severe qualitative impact on our major results, especially for all but the most recently formed hybrid zones.

@@

\subsection*{Islands of divergence}

\plr{seems like a nice point but should be more positive and less snarky.  "Thanks to studies like these, it is time to start explicit modeling, as we have done here..."}
The past years have witnessed evocative metaphors of seascapes to describe heterogeneous patterns of genetic divergence across the genome.  
The metaphors of ``islands'', ``continents'' and ``archipelagoes'' of divergence \citep{Turner2005, Nosil2009} have inspired much research (and controversy  \citep{Noor2009,Cruickshank2014}) in speciation genomics, but have been largely descriptive.  
REFS  http://www.ncbi.nlm.nih.gov/pubmed/16076241 http://www.ncbi.nlm.nih.gov/pubmed/19143936 http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2809014/ http://onlinelibrary.wiley.com/doi/10.1111/mec.12796/abstract
We hope that our work provides a path towards thinking about the process of heterogeneous divergence more concretely. % in terms of the processes of migration, and selection influencing patterns of ancestry across the genome of hybridizing populations.
Specifically, we provide a natural interpretation of large contiguous chunks of ancestry, as sites linked to those experiencing selection.   
Overall, focusing on lengths of ancestry blocks across the genome brings focus to the processes of migration and selection rather than high-level summaries that are somewhat abstracted from the evolutionary process.

@@

\subsection*{Simulation and analysis code}

These are written primarily in R, and are available HERE.

>>>>

\alisa{This process of a haplotype sampled in the present day tracing it's path $X_\tau$ backwards in time, while splitting and changing backgrounds $Z_\tau$ through recombination, is shown in Figure~\ref{Fig:schematic}.}
\plr{Connect these up more.}

@@

The simulations were executed in R, with scripts available on
\url{https://github.com/petrelharp/clinal-lineages}

@@

The above theory and simulations allow us to make predictions about patterns surrounding selected loci. In reality, however, such loci are not known, so it is useful to have per-site statistics that may allow for detection of candidate targets of selection. The most straightforward measure is $l_I(m)$,  the mean length of all haplotype chunks that contain genomic position $m$, given that $m$ is of ancestry $I$, where $I\in\{A,B\}$.
We then focus on $I_B(m)$ for $m$ across a chromosome. 
%Another statistic is the cumulative probability $C_I(m)$ of a sampled chunk encompassing position $m$ being longer than a chunk sampled from the distribution of all chunk lengths across the genome in the population.

Additionally, we look at the mean value of these statistics in the two chunks, $m-$ and $m+$ that flank, to the left and right respectively, the block of unbroken ancestry containing $m$. Since there appears to be a signal at these regions also, we further define a statistic

$$C(m) =  \frac{2\sum_i{l_B(m_i)}}{\sum_i{l_A(m_i+)+l_A(m_i-)}}$$

which is the ratio of mean block length around $m_i$ and mean length of blocks directly flanking the block surrounding $m_i$ (i.e. adjacent blocks).

We are interested in identifying regions of the genome with abnormal distributions of block lengths, and where indicated normalize $l_I(m)$ and $C(m)$ by dividing respectively by mean $l_I$ and $C$  within a deme. While there is useful information contained in the geographic distribution of block length patterns, obtaining good spatial sampling can often be difficult, and this allows us to predict patterns within relatively few populations.

Although our analytical model does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $T$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family, and is potentially a source of additional information. Within each deme $X$, we calculated $F_B(m,X)$, which is the number of independent ancestors from time $T$ contributing to the pool of alleles of ancestry $B$ at site $m$, divided by the number of individuals of ancestry $B$ at site $m$.

@@

As expected, regions surrounding a locus under selection are more resistant to introgression,
as seen in figures~\ref{Fig:resistanceToIntrogression1000g} and \ref{Fig:resistanceToIntrogression100g}).
When present, however, we expect haplotypes including the locally less common allele
to be longer than the genome-wide average.
Indeed,
as shown in Figures \ref{Fig:blockLengths} and \ref{Supp:blockLengthHeatmap},
the mean length of such haplotype blocks are up to three times longer
than the average for that geographic location.

% This pattern provides a potential avenue for identifying targets of selection by looking for a signal of increased $l_B$ or $l_A$ for positions surrounding a selected site, compared to the genome-wide background  at geographic sites far from the center of the zone (Figure~\ref{Fig:blockLengths},~\ref{Supp:blockLengthHeatmap}). 
The mean haplotype length found without conditioning on ancestry ($l_\cdot$)
also shows an small increase near the selected locus,
but it is much smaller (see Figures \ref{Supp:blockLengthHeatmapNoAnc} and \ref{Supp:blockLengthNoAnc}).

Additionally, we find that $l_A(m\pm)$, the length of blocks adjacent to blocks containing $m$, are shorter than average when $m$ is in a block of unbroken ancestry surrounding a selected site on the ``wrong side'' of the side of the zone.  (Figure~\ref{Supp:adjacentBlocks}). This pattern provides an avenue for further enhancement of the block-length signal around a selected locus, by measuring $C(m)$  (Figures~\ref{Supp:ratioBlockAdjacent},~\ref{Supp:ratioBlockAdjacentHeatmap}).

@@

@@

\begin{figure}
\includegraphics{figs/blocksAlongChromAncBConditioning}
\caption{Mean block length, $l_B(m_i)$, along a simulated chromosome of length $1M$ with selected locus of $s=0.01$ at position $0.5M$ in a hybrid zone of $T=1000$. Here each line represents  $l_B(\mathbf{k})$, the mean block lengths in a given deme, and is normalized by mean block length $l_B(m)$ along the chromosome in the deme.
    \plr{what is $T$ and $s$? also, add version with different $T$ and/or $s$ to the supplement.}
}\label{Fig:blockLengths}
\end{figure}

@@

\plr{This needs a bit more precise discussion.}
As expected, we find that $F_B(m)$ is highest on the size of the hybrid zone where individuals of ancestry $B$ predominate. We find that at the selected locus, $F_B(m,X)$ decreases with X, and is lowest far away from the zone center, where ancestry $B$ is at low frequency  (Figure~\ref{Fig:family_size}). As $m$ moves further away from the target of selection, the average family size recovers to expected values under neutrality.

One interpretation of this observation is that migrants tend to leave few offspring before being removed from the population by selection, leaving behind small families. In conjunction with block length, measuring $F$ can provide an additional signal for characterizing lineages strongly influenced by selection, by looking at levels of haplotype diversity in populations along a hybrid zone.

@@

\paragraph{The nature of selection}
Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models (e.g., ecological selection \citep{Barton1989,Barton1993}). 
We expect, therefore, our findings here to be somewhat generalizable.

@@

\paragraph{Geographic space and other sources of stochasticity}
For ease of calculations, our analytic model assumes a continuous and smooth one-dimensional spatial structure of the geographic landscape, allowing us to derive analytic expressions that track lineages. 
This assumes a large population density limit that excludes coalescence and makes lineages Markov.  
In contrast, our simulations model regularly spaced demes. 
In reality, populations may be patchily connected, especially at the edges of species ranges where hybrid zones may occur. 
While we still believe that the genomic signature of long ancestry chunks around selected sites will still hold in these situations, the strength of this expectation is unclear.

Of particular concern that stems from the above assumption is that our model ignores potential, and likely, sources of stochasticity arising from geographic structure and coalescence/pedigree structure. The latter can result in an underestimate of covariance in ancestry, as we have ignored additional sharing of ancestry through shared genealogy \cite{Liang2014}. With respect to genealogy, the broad agreement between our analytic predictions and simulations (which include an explicit pedigree structure), suggests that this assumption does not have a severe qualitative impact on our major results when populations are moderately high, especially for all but the most recently formed hybrid zones.

Lastly, while our theory focusses on a spatially arranged hybrid zone, it is likely that our results extend to alternative models of hybridization including models of assortative mating and hybrid swarms. This is because regardless of the specific demographic model, selection against hybrids will severely limit opportunities for recombination between ancestries around incompatibilities.

%, and is an example of the more general idea of duality for particle systems \alisa{I think this may be beyond the average ME reader?}. 
%\plr{Tracking lineages is duality, has nothing to do with 1D, though.}
%However, this assumes a large population density limit that excludes \plr{coalescence} and makes linages Markov \alisa{not sure about what point this is making? Is there a better way to state this for our intended audience?}.  
%\plr{It means we're ignoring drift/coalescence, an important source of stochasticity; but that's the next paragraph too}
%In contrast, our simulations model regularly spaced demes. In reality, populations may be patchily connected, especially at the edges of species ranges where hybrid zones may occur. While we still believe that the genomic signature of long ancestry chunks around selected sites will still hold in these situations, the strength of this expectation is unclear.

%Lastly, the analytic model presented here lacks \yb{both coalescence and a true pedigree structure}. \alisa{This can result in an underestimate of covariance in ancestry, as we have ignored sharing through genealogy \cite{Liang2014}.} 
%The broad agreement between our analytic predictions and simulations that include an explicit pedigree structure suggests that this assumption does not have a severe qualitative impact on our major results, especially for all but the most recently formed hybrid zones.

@@

@@

@@

\subsection*{Patterns of divergence}  
A number of studies have described heterogeneous patterns of genetic divergence across the genome. These so-called ``islands" of divergence \citep{Turner2005,Nosil2009} have been largely descriptive \citep{Cruickshank2014,Noor2009}, and as we begin to rigorously model patterns of divergence in the genome to begin to more  explore these patterns. The work presented here may offer ways in which to think about the process of heterogeneous divergence more concretely in terns of how processes such as migration and selection influence patterns across the genome of hybridizing populations. Specifically, we provide a natural interpretation of large contiguous chunks of ancestry, as sites linked to those experiencing selection.   
Overall, focusing on lengths of ancestry blocks across the genome brings focus to the processes of migration and selection rather than high-level summaries that are somewhat abstracted from the evolutionary process.

%\plr{seems like a nice point but should be more positive and less snarky.  "Thanks to studies like these, it is time to start explicit modeling, as we have done here..."}
%The past years have witnessed evocative metaphors of seascapes to describe heterogeneous patterns of genetic divergence across the genome.  
%The metaphors of ``islands'', ``continents'' and ``archipelagoes'' of divergence \citep{Turner2005, Nosil2009} have inspired much research (and controversy  \citep{Noor2009,Cruickshank2014}) in speciation genomics, but have been largely descriptive.  
%REFS  http://www.ncbi.nlm.nih.gov/pubmed/16076241 http://www.ncbi.nlm.nih.gov/pubmed/19143936 http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2809014/ http://onlinelibrary.wiley.com/doi/10.1111/mec.12796/abstract
%We hope that our work provides a path towards thinking about the process of heterogeneous divergence more concretely. % in terms of the processes of migration, and selection influencing patterns of ancestry across the genome of hybridizing populations.

@@

%\subsection*{Simulation and analysis code}
%These are written primarily in R, and are available HERE.
